\documentclass[twoside, numbers=noendperiod]{scrartcl}

\usepackage[blacklinks]{drangreport}

\usepackage{todonotes}
\usepackage{awlc-algebra}

\usetikzlibrary{calc,
				fadings,
				arrows,
				decorations.markings,
				decorations.pathreplacing,
        scopes}

\let\pp\relax
\let\tt\relax
\newcommand{\va}{\vec{a}}
\newcommand{\vb}{\vec{b}}
\newcommand{\vc}{\vec{c}}
\newcommand{\vv}{\vec{v}}
\newcommand{\ww}{\vec{w}}
\newcommand{\xx}{\vec{x}}
\newcommand{\yy}{\vec{y}}
\newcommand{\zz}{\vec{z}}
\newcommand{\pp}{\vec{p}}
\newcommand{\qq}{\vec{q}}
\newcommand{\rr}{\vec{r}}
\newcommand{\ee}{\vec{e}}
\newcommand{\tt}{\vec{t}}
\newcommand{\nn}{\vec{n}}

\newcommand{\ii}{\vec{i}}
\newcommand{\jj}{\vec{j}}
\newcommand{\kk}{\vec{k}}

\let\AA\relax
\newcommand{\AA}{\vec{A}}
\newcommand{\BB}{\vec{B}}
\newcommand{\CC}{\vec{C}}
\newcommand{\NN}{\vec{N}}

\newcommand{\HH}{\bb{H}}
\newcommand{\iH}{\in\HH}

\renewcommand{\vec}[1]{\mathbf{#1}}

\newcommand{\OO}{\vec{0}}

\newcommand{\zbar}{\overline{z}}
\newcommand{\wbar}{\overline{w}}

\DeclareMathOperator{\Area}{Area}
\DeclareMathOperator{\length}{length}
\DeclareMathOperator{\defect}{\delta}
\DeclareMathOperator{\Mobgp}{Mob}

\newcommand{\closure}[1]{\overline{#1}}
\newcommand{\boundary}[1]{\partial{#1}}

\renewcommand{\varphi}{\phi}

% Riemannian geometry
\newcommand{\ddif}[2]{\eval[0]{\dif{#1}}_{#2}}
\newcommand{\ssig}[1]{\eval[0]{\sigma_{#1}}_{\pp}}
\newcommand{\bsig}{\boldsymbol{\sigma}}

\renewcommand{\vec}[1]{#1}

\title{IB Geometry}

\newcommand\pgfmathsinandcos[3]{%
  \pgfmathsetmacro#1{sin(#3)}%
  \pgfmathsetmacro#2{cos(#3)}%
}
\newcommand\LongitudePlane[3][current plane]{%
  \pgfmathsinandcos\sinEl\cosEl{#2} % elevation
  \pgfmathsinandcos\sint\cost{#3} % azimuth
  \tikzset{#1/.estyle={cm={\cost,\sint*\sinEl,0,\cosEl,(0,0)}}}
}
\newcommand\LatitudePlane[3][current plane]{%
  \pgfmathsinandcos\sinEl\cosEl{#2} % elevation
  \pgfmathsinandcos\sint\cost{#3} % latitude
  \pgfmathsetmacro\yshift{\cosEl*\sint}
  \tikzset{#1/.estyle={cm={\cost,0,0,\cost*\sinEl,(0,\yshift)}}} %
}
\newcommand\DrawLongitudeCircle[2][1]{
  \LongitudePlane{\angEl}{#2}
  \tikzset{current plane/.prefix style={scale=#1}}
   % angle of "visibility"
  \pgfmathsetmacro\angVis{atan(sin(#2)*cos(\angEl)/sin(\angEl))} %
  \draw[current plane] (\angVis:1) arc (\angVis:\angVis+180:1);
  \draw[current plane,dashed] (\angVis-180:1) arc (\angVis-180:\angVis:1);
}
\newcommand\DrawLatitudeCircle[2][1]{
  \LatitudePlane{\angEl}{#2}
  \tikzset{current plane/.prefix style={scale=#1}}
  \pgfmathsetmacro\sinVis{sin(#2)/cos(#2)*sin(\angEl)/cos(\angEl)}
  % angle of "visibility"
  \pgfmathsetmacro\angVis{asin(min(1,max(\sinVis,-1)))}
  \draw[current plane] (\angVis:1) arc (\angVis:-\angVis-180:1);
  \draw[current plane,dashed] (180-\angVis:1) arc (180-\angVis:\angVis:1);
}

\tikzset{%
  % >=latex, % option for nice arrows
  % inner sep=0pt,%
  % outer sep=2pt,%
  mark coordinate/.style={inner sep=0pt,outer sep=0pt,minimum size=3pt,
    fill=black,circle}%
}

\begin{document}

\NotesTitle{IB}{Lent 2013}{Geometry}{Prof.~J.}{Rasmussen}
{
\section*{Course schedule}

Groups of rigid motions of Euclidean space. Rotation and reflection groups in two and three dimensions. Lengths of curves. \hfill [2]

Spherical geometry: spherical lines, spherical triangles and the Gauss-Bonnet theorem. Stereographic projection and Möbius transformations. \hfill [3]

Triangulations of the sphere and the torus, Euler number. \hfill [1]

Riemannian metrics on open subsets of the plane. The hyperbolic plane. Poincar\'{e} models and their metrics. The isometry group. Hyperbolic triangles and the Gauss-Bonnet theorem. The hyperboloid model. \hfill [4]

Embedded surfaces in $\R^3$. The first fundamental form. Length and area. Examples. \hfill [1]

Length and energy. Geodesics for general Riemannian metrics as stationary points of the energy. First variation of the energy and geodesics as solutions of the corresponding Euler-Lagrange equations. Geodesic polar coordinates (informal proof of existence). Surfaces of revolution. \hfill [2]

The second fundamental form and Gaussian curvature. For metrics of the form $\dif{u}^2 + G(u,v) \dif{v}^2$, expression of the curvature as $\sqrt{G_{uu}}/\sqrt{G}$\,. Abstract smooth surfaces and isometries. Euler numbers and statement of Gauss-Bonnet theorem, examples and applications. \hfill [3]

\subsection*{Appropriate books}

{\shortskip
P.M.H.~Wilson \emph{Curved Spaces}. CUP, January 2008 (£60 hardback, £24.99 paperback).

M.~Do Carmo \emph{Differential Geometry of Curves and Surfaces}. Prentice-Hall, Inc., Englewood Cliffs, N.J., 1976 (£42.99 hardback)

A.~Pressley \emph{Elementary Differential Geometry}. Springer Undergraduate Mathematics Series, Springer-Verlag London Ltd., 2001 (£19.00 paperback)

E.~Rees \emph{Notes on Geometry}. Springer, 1983 (£18.50 paperback)

M.~Reid and B.~Szendroi \emph{Geometry and Topology}. CUP, 2005 (£24.99 paperback)}}

\TableofContents

\input{geo-chap-01}		% Euclidean geometry
\input{geo-chap-02}		% Spherical geometry
\input{geo-chap-03}		% Möbius transformations
\input{geo-chap-04}   % Riemannian geometry
\input{geo-chap-05}   % Hyperbolic geometry
\input{geo-chap-06}   % Geodesics
\input{geo-chap-07}   % Surfaces

\input{geo-chap-a1}   % Review sheets

\lecturenotesendshort

\end{document}