\documentclass[twoside, numbers=noendperiod]{scrartcl}

\usepackage[blacklinks]{drangreport}

\usepackage{awlc-algebra}
\usepackage{multirow}

\newcommand{\xx}{\vec{x}}
\newcommand{\yy}{\vec{y}}

\newcommand{\closure}[1]{\overline{#1}}

\DeclareMathOperator{\Int}{Int}
\DeclareMathOperator{\Cl}{Cl}

\newcommand{\Eucl}{\text{Eucl}}

\newcommand{\Abar}{\overline{A}}
\newcommand{\fbar}{\overline{f}}

\newcommand{\xbar}{\overline{x}}
\newcommand{\ybar}{\overline{y}}

\newcommand{\boundary}{\del}

\newcommand{\base}{\cal{B}}
\newcommand{\cover}{\cal{U}}
\newcommand{\subcover}{\cal{V}}

\title{IB Metric \& Topological Spaces}

\begin{document}

\NotesTitle{IB}{Easter 2012}{Metric \& Topological Spaces}{Prof.~P.~M.~H.}{Wilson}
{
\section*{Course schedule} % (fold)
\label{sub:course_schedule}

\subsubsection*{Metrics} % (fold)
\label{ssub:metrics}

Definition and examples. Limits and continuity. Open sets and neighbourhoods. Characterizing limits and continuity using neighbourhoods and open sets. \hfill [3]

% subsubsection metrics (end)

\subsubsection*{Topology} % (fold)
\label{ssub:topology}

Definition of a topology. Metric topologies. Further examples. Neighbourhoods, closed sets, convergence and continuity. Hausdorff spaces. Homeomorphisms. Topological and non-topological properties. Completeness. Subspace, quotient and product topologies. \\
\mbox{  } \hfill [3]

% subsubsection topology (end)

\subsubsection*{Connectedness} % (fold)
\label{ssub:connectedness}

Definition using open sets and integer-valued functions. Examples, including intervals. Components. The continuous image of a connected space is connected. Path-connectedness. Path-connected spaces are connected but not conversely. Connected open sets in Euclidean space are path-connected. \hfill [3]

% subsubsection connectedness (end)

\subsubsection*{Compactness} % (fold)
\label{ssub:compactness}

Definition using open covers. Examples: finite sets and $[0,1]$. Closed subsets of compact spaces are compact. Compact subsets of a Hausdorff space must be closed. The compact subsets of the real line. Continuous images of compact sets are compact. Quotient spaces. Continuous real-valued functions on a compact space are bounded and attain their bounds. The product of two compact spaces is compact. The compact subsets of Euclidean space. Sequential compactness. \hfill [3]

% subsubsection compactness (end)

% subsubsection modules (end)

% subsection course_schedule (end)

\subsection*{Appropriate books} % (fold)
\label{sub:literature}

{\shortskip
W.A.~Sutherland \emph{Introduction to Metric and Topological Spaces}. Clarendon 1975 (£21.00 paperback).

A.J.~White \emph{Real Analysis: an Introduction}. Addison-Wesley 1968 (out of print)

B.~Mendelson \emph{Introduction to Topology}. Dover, 1990 (£5.27 paperback)}

% subsection literature (end)
}

\TableofContents

\input{mettop-chap-01}		% Metric spaces
\input{mettop-chap-02}		% Topological spaces
\input{mettop-chap-03}		% Connectedness
\input{mettop-chap-04}		% Compactness

\lecturenotesend

\end{document}